Global search algorithms for one-dimensional multiextremal optimization problems with nonlinear constraints

### 2017-10-04 2017/ParOptBook/Chapter4.tex
Some misprints have been corrected

\section{Problem statement}
In the present chapter, the one-dimensional problems of the constrained optimization will
be considered. Let $\varphi(x)$ and $g_j(x)\le 0,1\le j\le m$ be the real functions defined in a closed interval $[a,b]$ of the real axis. It is requires to find the values $x^*$ satisfying the condition
\begin{equation}
  \label{eq4:problem}
  \varphi(x^*)=\min\{\varphi(x):x\in [a,b], g_j(x)\le 0,\; 1\le j\le m\}
\end{equation}

As before, the objective function $\varphi$ (hereinafter denoted also as $g_{m+1}$) and the left parts of the constraints $g_j(x)\le 0,1\le j\le m$ re assumed to satisfy \emph{Lipschitz condition} with the corresponding constants $L_j,1\le j\le m+1$, namely:
\begin{equation}
  \label{eq4:lip_condition}
  |g_j(x_1)-g_j(x_2)|\le L_j|x_1-x_2|,1\le j\le m+1,x_1,x_2\in [a,b]
\end{equation}

\example{
\label{ex4:problem}
Let us consider a problem of type \eqref{eq4:problem} where $x\in [0.6;2.2],m=2$,
\begin{gather*}
  \varphi(x)=\cos(18x-3)\sin(10x-7)+1.5, \\
  g_1(x)=\exp(-x/2)\sin(6x-1.5), \\
  g_2(x)=|x|\sin(2\pi x-0.5).
\end{gather*}
he exact solution of the problem is $x^*=2.0795,\varphi(x^*)=0.565$. The graphs of the functions $g_1(x),g_2(x)$,and $\varphi(x)$ are presented in Fig. \ref{fig:4_1}.
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.8\textwidth]{figures/4_1.png}
  \caption{One-dimensional global optimization problem}
  \label{fig:4_1}
\end{figure}
}

\section{Partial computability and index scheme of the account of constraints}
Problem \eqref{eq4:problem} can be considered in the statement when each function $g_j,1\le j\le m+1$ is defined and computable only in the corresponding subdomains $Q_j\subset [a,b]$ where
\begin{equation}
  \label{eq4:q}
  Q_1=[a,b],Q_(j+1)=\{x\in Q_j:g_j(x)\le 0\},1\le j\le m.
\end{equation}

@@

For the purposes of further treatment, let us introduce a classification of the points $x$ in the search domain $[a,b]$ using an index $\nu=\nu(x)$ where $\nu-1$ is the number of constraints, which are fulfilled in this point. Formally, the index $\nu$ is defined by the conditions
\begin{equation}
  \label{eq4:condition}
  g_j(x)\le 0,1\le j \le \nu-1,g_\nu(x)>0,
\end{equation}
(the last inequality is inessential  if $\nu=m+1$) and it satisfies the inequalities
\begin{displaymath}
  1\le\nu=\nu(x)\le m+1.
\end{displaymath}
The problem of example \ref{ex4:problem} assuming a partial computability of the functions is presented in Fig. \ref{fig:4_2}. The arcs of the restrictions $g_1(x),\:g_2(x)$, and of the objective function $\varphi(x)$ defined in the corresponding subdomains $Q_j$ from \eqref{eq4:q} are shown. The point $x^1$ with the index $\nu(x^1)=1$, the point $x^2$ with the index $\nu(x^2)=2$, and the point $x^3$ with the index $\nu(x^3)=3$ are shown also.

@@

\section{The index global search algorithm}
The first trial is executed at an arbitrary internal point $x_1\in(a,b)$. The selection of any subsequent trial point $x^{k+1},k\ge 1$ is determined by the following rules.

@@

Let us formulate the conditions of convergence for the algorithm in the form of the following theorem.
\begin{theorem}
Let the considered index algorithm be applied to the solving of problem \eqref{eq4:problem} and the following conditions be satisfied:
\begin{enumerate}
  \item each function $gj, 1\le j\le m+1$ satisfies Lipschitz condition with the constant $L_j$ in the interval $[a,b]$, i.e.,
  \[
  |g_j(x_1)-g_j(x_2)|\le L_j|x_1-x_2|,1\le j\le m+1,x_1,x_2 \in[a,b]
  \]
  \item the following inequalities hold for the quantities $\mu_\nu$ from \eqref{eq4:lip_est} starting from some step
  \[
  r_\nu\mu_\nu > 2L_\nu,1\le\nu\\le m+1.
  \]
\end{enumerate}

@@

n the cases, when it is necessary to denote the parameters $m$ and $\delta$ used by the generator during its functioning, let us use the notation $G ( m , \delta )$. In the experiments, the results of whichare presented below two generators were employed. One of these (let us denote it as $G_{SH} ( 5 , 1 ) )$ was based on the function
\begin{displaymath}
  \varphi(x)=-\sum_{j=1}^{10}\frac{1}{(K_j(x-A_j)^2+C_j)}x\in[0,10].
\end{displaymath}
$1\le K_j\le 3,\: 0\le A_j,\: C_j\le 10 $, where the parameters $K_j,\: A_j,\: C_j ,\: 1 \le j \le 10$ are the independent random variables distributed uniformly over the intervals specified above. The second generator hereinafter denoted as $G_{HL} ( 5 , 1 ) )$ was based on the function
\begin{equation}
  \varphi(x)=\sum_{j=1}^{14}(A_j\sin(2j\pi x) + B_j\cos(2j\pi x)),\: x\in[0,1]
\end{equation}
where the values of the parameters $A_j,\: B_j,\: 1 \le j \le 14$, are distributed independently and uniformly over the interval $[-1 ,1 ]$. Both generators were applied to the generation of 100 own
problems of the kind \eqref{eq4:problem}.

>>>>

\section{Problem statement}
In the present chapter, the one-dimensional problems of the constrained optimization will
be considered. Let $\varphi(x)$ and $g_j(x)\le 0,1\le j\le m$, be the real functions defined in a closed interval $[a,b]$ of the real axis. It is requires to find the values $x^*$ satisfying the condition
\begin{equation}
  \label{eq4:problem}
  \varphi(x^*)=\min\{\varphi(x):x\in [a,b], g_j(x)\le 0,\; 1\le j\le m\}.
\end{equation}

As before, the objective function $\varphi$ (hereinafter denoted also as $g_{m+1}$) and the left parts of the constraints $g_j(x)\le 0,1\le j\le m$, are assumed to satisfy \emph{Lipschitz condition} with the corresponding constants $L_j,1\le j\le m+1$, namely:
\begin{equation}
  \label{eq4:lip_condition}
  |g_j(x_1)-g_j(x_2)|\le L_j|x_1-x_2|,1\le j\le m+1,x_1,x_2\in [a,b].
\end{equation}

\example{
\label{ex4:problem}
Let us consider a problem of type \eqref{eq4:problem} where $x\in [0.6,2.2],\; m=2$,
\begin{gather*}
  \varphi(x)=\cos(18x-3)\sin(10x-7)+1.5, \\
  g_1(x)=\exp(-x/2)\sin(6x-1.5), \\
  g_2(x)=|x|\sin(2\pi x-0.5).
\end{gather*}
The exact solution of the problem is $x^*=2.0795,\; \varphi(x^*)=0.565$. The graphs of the functions $g_1(x),g_2(x)$ and $\varphi(x)$ are presented in Fig. \ref{fig:4_1}.
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.8\textwidth]{figures/4_1.png}
  \caption{One-dimensional global optimization problem}
  \label{fig:4_1}
\end{figure}
}

\section{Partial computability and index scheme of the account of constraints}
Problem \eqref{eq4:problem} can be considered in the statement when each function $g_j,1\le j\le m+1$, is defined and computable only in the corresponding subdomains $Q_j\subset [a,b]$, where
\begin{equation}
  \label{eq4:q}
  Q_1=[a,b],Q_(j+1)=\{x\in Q_j:g_j(x)\le 0\},1\le j\le m.
\end{equation}

@@

For the purposes of further treatment, let us introduce a classification of the points $x$ in the search domain $[a,b]$ using an index $\nu=\nu(x)$ where $\nu-1$ is the number of constraints, which are fulfilled in this point. Formally, the index $\nu$ is defined by the conditions
\begin{equation}
  \label{eq4:condition}
  g_j(x)\le 0,\; 1\le j \le \nu-1,\; g_\nu(x)>0,
\end{equation}
(the last inequality is inessential  if $\nu=m+1$) and it satisfies the inequalities
\begin{displaymath}
  1\le\nu=\nu(x)\le m+1.
\end{displaymath}
The problem of example \ref{ex4:problem} assuming a partial computability of the functions is presented in Fig. \ref{fig:4_2}. The arcs of the restrictions $g_1(x),\:g_2(x)$, and of the objective function $\varphi(x)$ defined in the corresponding subdomains $Q_j$ from \eqref{eq4:q} are shown. The point $x^1$ with the index $\nu(x^1)=1$, the point $x^2$ with the index $\nu(x^2)=2$, and the point $x^3$ with the index $\nu(x^3)=3$ are shown also.

@@

\section{The index global search algorithm}
The first trial is executed at an arbitrary internal point $x_1\in(a,b)$. The selection of any subsequent trial point $x^{k+1},\; k\ge 1$ is determined by the following rules.

@@

Let us formulate the conditions of convergence for the algorithm in the form of the following theorem.
\begin{theorem}
Let the considered index algorithm be applied to the solving of problem \eqref{eq4:problem} and the following conditions be satisfied:
\begin{enumerate}
  \item each function $gj, 1\le j\le m+1$ satisfies Lipschitz condition with the constant $L_j$ in the interval $[a,b]$, i.e.,
  \[
  |g_j(x_1)-g_j(x_2)|\le L_j|x_1-x_2|,\; 1\le j\le m+1,\; x_1,x_2 \in[a,b]
  \]
  \item the following inequalities hold for the quantities $\mu_\nu$ from \eqref{eq4:lip_est} starting from some step
  \[
  r_\nu\mu_\nu > 2L_\nu,\; 1\le\nu \leq m+1.
  \]
\end{enumerate}

@@

In the cases, when it is necessary to denote the parameters $m$ and $\delta$ used by the generator during its functioning, let us use the notation $G ( m , \delta )$. In the experiments, the results of whichare presented below two generators were employed. One of these (let us denote it as $G_{SH} ( 5 , 1 ) )$ was based on the function
\begin{displaymath}
  \varphi(x)=-\sum_{j=1}^{10}\frac{1}{(K_j(x-A_j)^2+C_j)},\; x\in[0,10].
\end{displaymath}
$1\le K_j\le 3,\: 0\le A_j,\: C_j\le 10 $, where the parameters $K_j,\: A_j,\: C_j ,\: 1 \le j \le 10$ are the independent random variables distributed uniformly over the intervals specified above. The second generator hereinafter denoted as $G_{HL} ( 5 , 1 ) )$ was based on the function
\begin{equation}
  \varphi(x)=\sum_{j=1}^{14}(A_j\sin(2j\pi x) + B_j\cos(2j\pi x)),\: x\in[0,1]
\end{equation}
where the values of the parameters $A_j,\: B_j,\: 1 \le j \le 14$, are distributed independently and uniformly over the interval $[-1 ,1 ]$. Both generators were applied to the generation of 100 own
problems of the kind \eqref{eq4:problem}.
